--- 
slug: vignettes 
title: "Optimal workflows for package vignettes" 
authors: 
- Maëlle Salmon 
date: "2020-06-03" 
tags: 
- package development 
- documentation
output: hugodown::hugo_document
---

```{r setup, include=FALSE}
library("printr")

# Options to have images saved in the post folder
# And to disable symbols before output
knitr::opts_chunk$set(fig.path = "", comment = "")

# knitr hook to make images output use Hugo options
knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)

# knitr hook to use Hugo highlighting options
knitr::knit_hooks$set(
  source = function(x, options) {
  hlopts <- options$hlopts
    paste0(
      "```r ",
      if (!is.null(hlopts)) {
      paste0("{",
        glue::glue_collapse(
          glue::glue('{names(hlopts)}={hlopts}'),
          sep = ","
        ), "}"
        )
      },
      "\n", glue::glue_collapse(x, sep = "\n"), "\n```\n"
    )
  }
)
```

Yet another post with a focus on [package documentation](/tags/documentation/)!
This time, we'll cover vignettes a.k.a "long-form package documentation".
What are they? Where do they live?
How do you create one without upsetting R CMD check?
How do you ensure user-friendliness?

## What is a vignette? Where does it live?

In this section we shall go over basics of package vignettes.

### Vignette 101

In the "R packages" book by Hadley Wickham and Jenny Bryan, the [vignettes chapter](https://r-pkgs.org/vignettes.html) starts with _"A vignette is a long-form guide to your package. Function documentation is great if you know the name of the function you need, but it’s useless otherwise."_[^disagree]
In ["Writing R Extensions"](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Writing-package-vignettes), vignettes are defined as _"documents in PDF or HTML format obtained from plain-text literate source files from which R knows how to extract R code and create output (in PDF/HTML or intermediate LaTeX)."_.

In practice, if your package contains one or several vignette(s), an user could 

* find them using the `vignette()` or `browseVignettes()` function, for instance they could type `vignette(package = "rhub")` or `browseVignettes(package = "rhub")` to access the list of installed vignettes for the `rhub` package[^printr]

```{r rhub-vignette-list}
vignette(package = "rhub")
```

```{r rhub-vignette-list2}
browseVignettes("rhub")
```

* see them from the [CRAN page of the package](https://cran.r-project.org/web/packages/rhub/index.html), and its [`pkgdown` website](https://r-hub.github.io/rhub/) if there's one.

As a package author you could be fine only knowing about [`usethis::use_vignette()`](https://usethis.r-lib.org/reference/use_vignette.html) for creating a vignette, and that packages used in the vignette need to be listed in `DESCRIPTION` (under `Suggests` if they're only used in the vignette).
Still, it's useful to know about vignettes for debugging problems or finding
workarounds for issues you might encounter.

### Infrastructure for vignettes

The building of package vignettes can either use the default Sweave vignette engine, or [a vignette engine provided by a CRAN package](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Non_002dSweave-vignettes) like [`knitr` by Yihui Xie](https://bookdown.org/yihui/rmarkdown-cookbook/package-vignette.html).
[`knitr::rmarkdown` vignette engine](https://community.rstudio.com/t/question-about-usethis-vignette-template/32048) is the one recommended in the R packages book, and `usethis`.
It allows writing vignettes in R Markdown.

[See the source of `rhub` main vignette](https://github.com/r-hub/rhub/blob/master/vignettes/rhub.Rmd). 
It has YAML metadata at the top, some non-executed code chunks, some executed code chunks.
To allow for that vignette to be built, a [field in `DESCRIPTION`](https://github.com/r-hub/rhub/blob/6ae6f35e958f3beab1e2c8e6f704affa23c8ce29/DESCRIPTION#L47) mentions the vignette engine:

```yaml
VignetteBuilder: knitr
```

The boilerplate Rmd under a new `vignettes` folder, and that `DESCRIPTION` field, are what `usethis::use_vignette()` would create for you.
Then you can write as you would a standard R Markdown document, knitting for previewing it.

Other vignette builders include [`R.rsp`](https://cran.r-project.org/web/packages/R.rsp/index.html) that we'll mention again later, [`noweb`](https://cran.r-project.org/web/packages/noweb/index.html) to use the [noweb literate programming tool](https://en.wikipedia.org/wiki/Noweb) (which actually looks a lot like sweave?), [`rasciidocs`](https://cran.r-project.org/web/packages/rasciidoc/index.html) that was recently archived at the time of writing.
It is unlikely you'll want to write your own vignette engine. 

How many packages use a non-Sweave vignette?
One way to assess that is to look for packages that have a `VignetteBuilder` field in `DESCRIPTION` with R-hub's own [`pkgsearch`](http://r-hub.github.io/pkgsearch/).[^twovb]

```{r count-vignettebuilder}
results <- pkgsearch::advanced_search("_exists_" = "VignetteBuilder")
attr(results, "metadata")$total

knitr <- pkgsearch::advanced_search(VignetteBuilder = "knitr")
attr(knitr, "metadata")$total
# for comparison
nrow(available.packages())
```

Quite a lot, about `r round(attr(results, "metadata")$total/nrow(available.packages()), digits = 2)*100`% of CRAN pages use a non Sweave vignette engine and about `r round(attr(knitr, "metadata")$total/nrow(available.packages()), digits = 2)*100`% use knitr for at least one vignette![^twovb]
Other packages might have _Sweave_ vignettes, and some CRAN packages don't have vignettes, whereas having a vignette is compulsory for Bioconductor packages.

### Overview of vignettes states

Following the [neat diagram of the R packages book](https://r-pkgs.org/package-structure-state.html),

* You write your vignette(s) in the `vignettes/` folder. [`rhub` source](https://github.com/r-hub/rhub).

* During building vignettes are [built](https://github.com/wch/r-source/blob/1d4f7aa1dac427ea2213d1f7cd7b5c16e896af22/src/library/tools/R/build.R#L320) and then vignettes sources, outputs, and anything written in [`install_extras`](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#index-_002einstall_005fextras-file) (a friend of [`.Rbuildignore` and `.Rinstignore`](/2020/05/20/rbuildignore/) except it shows what to _keep_ not _discard_!) gets moved to `inst/doc/`. [`rhub` contents on CRAN](https://github.com/cran/rhub). 
 
:bulb: If your vignette shows an external image not generated by the build process, you also need to include it in `install_extras`, 

* During installation the content of `inst/doc/` get copied to `doc/`. `rhub` content in my local library:

```{r rhub}
fs::dir_tree(find.package("rhub"))
```

### Your vignette for R CMD check

So, sometimes R CMD check will throw errors related to vignette building. 
How to deal with them?

:bulb: There is [good troubleshooting advice in the R packages book](https://r-pkgs.org/vignettes.html#vignette-cran)

:bulb: Based on what we said in the previous subsection, 
CMD build builds vignettes from `vignettes/`  
whereas R CMD check checks they can be rebuilt from `inst/doc/`. 
So if there were data in `vignettes/`, given it's not copied to `inst/doc/`... 
R CMD check will error!

It's also useful to know that there are options related to vignette building and checking in R CMD build and R CMD check.
Of course you don't control these options for CRAN but you do control them when sending your packages to R-hub package builder, and when setting up continuous integration.
See for instance [this great tip by John Blischak](https://community.rstudio.com/t/compute-intensive-vignettes-devtools-and-travis-ci/45865/6), _"checking the package while ignoring the vignettes can be done with the following steps:"_

```r
R CMD build --no-build-vignettes --no-manual .
R CMD check --no-manual --ignore-vignettes --as-cran *. tar.gz
```

For R-hub package builder,

* To tweak the build you need to build your package yourself (from the command line or with `devtools::build()`) and indicate the path to the tarball, as opposed to the package source, in your call to [`rhub::check()`](https://r-hub.github.io/rhub/reference/check.html)

* You can tweak the R CMD check by using the `check_args` argument.

## Workaround workflows for vignettes

In this section we'll go over workarounds for some common vignette "problems".

### How to include my pre-print / cheatsheet as a PDF vignette?

Sometimes you might want to include a PDF as a vignette, without wanting to deal with missing LaTeX dependencies; or because the PDF is not knit from R (a cheatsheet); or the computations are too long. 
In that case there are two alternatives:

* Following the process described in [a blog post by 
Mark van der Loo, entitled _"Add a static pdf vignette to an R package"_](http://www.markvanderloo.eu/yaRb/2019/01/11/add-a-static-pdf-vignette-to-an-r-package/);

* Using the `R.rsp` package by Henrik Bengtsson.

As an example of `R.rsp` usage, the [`treeBUGS` package](https://cran.r-project.org/web/packages/TreeBUGS/) has [HTML vignettes](https://cran.r-project.org/web/packages/TreeBUGS/vignettes/TreeBUGS_1_intro.html), and a [PDF vignette corresponding to a pre-print](https://cran.r-project.org/web/packages/TreeBUGS/vignettes/Heck_2018_BRM.pdf).
In its [`DESCRIPTION`](https://github.com/danheck/TreeBUGS/blob/9983dd0597717950557f3dc4ccaa7b118b24d864/DESCRIPTION#L35) it indicates `R.rsp` as one of the vignette engines.

```yaml
VignetteBuilder: 
	knitr,
	R.rsp
```

In the `vignettes/` folder of its source one sees [a file called `Heck_2018_BRM.pdf.asis`](https://github.com/danheck/TreeBUGS/blob/master/vignettes/Heck_2018_BRM.pdf.asis)

```tex
%\VignetteIndexEntry{Heck, Arnold, & Arnold (2018): TreeBUGS paper (Behavior Research Methods)}
%\VignetteEngine{R.rsp::asis}
%\VignetteKeyword{PDF}
%\VignetteKeyword{HTML}
%\VignetteKeyword{vignette}
%\VignetteKeyword{package}
%\VignetteKeyword{TreeBUGS}
```

### How to include a compute-intensive / authentication-dependent vignette?

A very similar problem can happen with HTML vignettes, when their computations are too long, or depend on a system dependency or authentication token absent from CRAN machines -- hence R CMD check would fail for sure.
So, what can you do?

* You could [pre-compute vignettes following the approach described by Jeroen Ooms in an rOpenSci tech note](https://ropensci.org/technotes/2019/12/08/precompute-vignettes/). The gist is to call actual Rmd vignette source something like `.Rmd.orig` and to knit them to .Rmd. The .Rmd fake vignette sources have already executed R code. Therefore it can be used in the R CMD build/check process without creating errors, it can be knit rapidly.

* You could indeed use `purl` & `eval`, as [global knitr options](https://community.rstudio.com/t/precompiling-vignette-with-devtools/1583/6). See for instance [this GitHub thread](https://github.com/hrecht/censusapi/issues/32). A chunk could be line [the one from `googlesheet`](https://github.com/jennybc/googlesheets/blob/master/vignettes/managing-auth-tokens.Rmd#L15-L23)

```Markdown
```{r, echo = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  purl = NOT_CRAN,
  eval = NOT_CRAN
)
```
````

* You could skip having vignettes and make them "articles" instead, that are present in a `pkgdown` site but not on CRAN/Bioconductor. [`googledrive` setup](https://github.com/tidyverse/googledrive/tree/master/vignettes). Of course it means vignettes are not available for offline consumption.

In the two previous subsections we recommended pre-building stuff, which might make some people cringe, but we like this [quote by Henrik Bengtsson in R-package-devel](https://www.mail-archive.com/r-package-devel@r-project.org/msg00812.html).

> Some may argue that you're package is not fully tested this way, but that depends on how well your package tests/ are written. I tend to look at examples() and vignettes as demos, and tests/ as actually tests. All should of course pass R CMD check and run, but the tests/ are what really test the package.

He also makes the point,

>  For
reproducibility, I would include the root/source vignette in the
package as well, e.g. in inst/full-vignettes/ with instructions and/or
a function on how to rebuild it.


## User-friendly vignettes

How to make vignettes pretty, and information easy to find?

### Pretty vignettes

You might want to tweak layout and aspect of your vignette a bit, maybe with [custom CSS](https://bookdown.org/yihui/rmarkdown/r-package-vignette.html)[^bc].
Using a [disappointingly unspecific GitHub code search on R-hub mirror of CRAN](https://github.com/search?l=&o=desc&q=css+user%3Acran++extension%3ARmd+path%3Ainst%2Fdoc&s=indexed&type=Code) we found the example of [`idiogramFISH`](https://gitlab.com/ferroao/idiogramFISH/-/tree/master) that [defines](https://gitlab.com/ferroao/idiogramFISH/-/tree/master/vignettes/css) and [uses](https://gitlab.com/ferroao/idiogramFISH/-/blob/master/vignettes/AplotIdiogramsVig.Rmd#L33) custom stylesheets for its vignette, that makes the vignette look very modern [on its CRAN page](https://cran.r-project.org/web/packages/idiogramFISH/vignettes/AplotIdiogramsVig.html)!
Note that it also uses some JavaScript for the table of content and "return to top" links, definitely not light-weight styling.

We haven't tried this ourselves with the latest RStudio IDE version but [apparently there can be conflicts between your vignette background and RStudio IDE theme](https://community.rstudio.com/t/html-vignettes-with-dark-themed-ide/3532).
Any experience report?

Now, an even better way to tweak your vignettes is to invest some time in creating a `pkgdown` website that will feature both manual pages, vignettes, changelogs, etc.
Should your local vignettes contain a link to the `pkgdown` version to be sure that users that look at an offline vignette but have an internet connection can get a better user experience?
And regarding the offline experience, would it make sense to also generate a PDF version of HTML vignettes, maybe with paged.js[^pagejs]?

### Cross-references

Link from vignettes to docs, from docs to vignettes.
Link pkgdown websites from vignettes?


 vignettes and pkgdown When
mentioning pkgdown, special role of the vignette pkgname.Rmd + mention tweaking
the navbar. Articles index. Recommended workflows for standard stuff. Rmd
fragments Workflows for working around issues (prebuilding Jeroen's note, PDF
stuff; purl&eval; articles not vignettes) Further stuff: PDF with paged.js?
Vignettes or book? See drake

### Repeat yourself

Even better than cross-references, or complementary to them is the idea to repeat yourself.
As a quick reminder from [our post about READMEs](/2019/12/03/readmes/#tools-for-writing-and-re-using-content), and as [explained very well by Garrick Aden-Buie](https://www.garrickadenbuie.com/blog/dry-vignette-and-readme/), you can re-use Rmd fragments in your package README, vignettes and manual pages without actually needing to copy-paste content!

## Conclusion

In this post we offered a quite detailed, but probably not exhaustive, guide around R package vignettes.
We didn't really discuss [_content_ of vignettes](https://r-pkgs.org/vignettes.html#vignette-advice), how to best assess their usefulness (surveys? traffic data?), or their use as a way to [encapsulate analyses in a package structure or "research compendium"](https://annakrystalli.me/rrresearchACCE20/creating-a-research-compendium-with-rrtools.html).
Do you have any special vignette setup or favorite trick?
Don't hesitate to share!

[^printr]: For rendering the vignettes list in this post we used the [`printr`](https://yihui.org/printr/#vignette-dataset-lists) package.
[^bc]: Bioconductor has its [own vignette style](https://www.bioconductor.org/packages/release/bioc/html/BiocStyle.html).
[^pagejs]: Compared with [`pagedown`](https://pagedown.rbind.io/) the vague idea mentioned here would mean adding a custom print stylesheet to the vignette and using the [paged.js CLI](https://gitlab.pagedmedia.org/tools/pagedjs-cli) for generating a PDF locally before submission, PDF that'd be present in `inst/doc`, and linked to from the html vignette.
[^disagree]: Note that on a `pkgdown` website, a [well-organized reference](https://pkgdown.r-lib.org/articles/pkgdown.html#reference-1) page can help make function documentation more useful.
[^twovb]: A query like `pkgsearch::advanced_search(VignetteBuilder = "knitr AND R.rsp")` would show how many packages use both `knitr` and `R.rsp` as vignette engines, meaning they have at least one vignette using `knitr` and one vignette using `R.rsp`.