--- 
slug: vignettes 
title: "Optimal workflows for package vignettes" 
authors: 
- Maëlle Salmon 
date: "2020-06-03" 
tags: 
- package development 
- documentation
output: hugodown::hugo_document
---

```{r setup, include=FALSE}
library("printr")

options(knitr.table.format = "markdown")

# Options to have images saved in the post folder
# And to disable symbols before output
knitr::opts_chunk$set(fig.path = "", comment = "")

# knitr hook to make images output use Hugo options
knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)

# knitr hook to use Hugo highlighting options
knitr::knit_hooks$set(
  source = function(x, options) {
  hlopts <- options$hlopts
    paste0(
      "```r ",
      if (!is.null(hlopts)) {
      paste0("{",
        glue::glue_collapse(
          glue::glue('{names(hlopts)}={hlopts}'),
          sep = ","
        ), "}"
        )
      },
      "\n", glue::glue_collapse(x, sep = "\n"), "\n```\n"
    )
  }
)
```

Yet another post with a focus on [package documentation](/tags/documentation/)!
This time, we'll cover vignettes a.k.a "long-form package documentation",
showing how your package functionalities work together and within the larger ( R
) ecosystem.
## What is a vignette? Where does it live?

Etymology. other languages e.g. Python?

### Vignette 101

What's a vignette (goal see R package book, in practice WRE) 
Where does the user see them
Now you could be fine only knowing about `usethis::use_vignette()` but Say it's useful to know about it for debugging problems or finding
workarounds (e.g. prebuilt vignettes, installextras).

An user could type `vignette(package = "rhub")` or `browseVignettes(package = "rhub")` to access the list of installed vignettes for the `rhub` package[^printr]

```{r rhub-vignette-list}
vignette(package = "rhub")
```

```{r rhub-vignette-list2}
browseVignettes("rhub")
```



### Infrastructure for vignettes

vignette builder. dependencies.

### Overview of vignettes states

Where is the
vignette for the user, where is it the package source/tarball, how is it copied.
installation from gitHub

### Your vignette for R CMD check

 R CMD check
options around vignettes, reminders of how to pass them to R-hub package
builder. 

## Workflows around vignettes

### How to include my pre-print / cheatsheet as a vignette?

why, indexing.

### How to include a compute-intensive / authentication-dependent vignette?

my post on RStudio community

### Re-use content from the README and manual pages

Rmd fragments

## User-friendly vignettes

How to make vignettes pretty, and information easy to find?

### Pretty vignettes

pkgdown

LaTeX styles, CSS -- careful about browser vs RStudio.

### Cross-references

Link from vignettes to docs, from docs to vignettes.
Link pkgdown websites from vignettes?


 vignettes and pkgdown When
mentioning pkgdown, special role of the vignette pkgname.Rmd + mention tweaking
the navbar. Articles index. Recommended workflows for standard stuff. Rmd
fragments Workflows for working around issues (prebuilding Jeroen's note, PDF
stuff; purl&eval; articles not vignettes) Further stuff: PDF with paged.js?
Vignettes or book? See drake

[^printr]: For rendering the vignettes list in this post we used the [`printr`](https://yihui.org/printr/#vignette-dataset-lists) package.