---
slug: cliff-notes-about-cli
title: "Cliff notes about the cli package" 
authors: 
- Maëlle Salmon
- Athanasia Mo Mowinckel
date: "2023-10-20" 
tags: 
- package development
- code style
output: hugodown::hugo_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.path = "", comment = "")
# knitr hook to make images output use Hugo options
knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)

```

We've both coincidentally dipped our toes in the wonderful world of pretty messaging with [cli](https://cli.r-lib.org/).

```{r, eval=FALSE}
cli::cli_li("marvellous package adopted!")
```

In this post, we transform the hurdles we encountered in a series of tips so that your own journey might be slightly smoother, and also to encourage you to try out cli in your package!

Note that blogging about an interface-building tool is a bit lame because the output is pretty... in a console, not in a Markdown file!
So don't hesitate to open an R console to see it for yourself.

## cli is the thing for package interfaces now!

You can view cli as a domain-specific language (DSL) for command-line interfaces (CLI): Just like tidyverse makes your data pipelines easier to construct and more readable, cli makes your communication producing code simpler to write!

cli, well its author Gábor Csárdi, deals with pesky details and we package developers only need to use the high-level interface.
As an example, text is automatically wrapped to the terminal width!

cli is truly feature-rich, for instance allowing you to make URLs in messages clickable and code in messages runnable at a click!

cli is part of the [tidyverse capsule wardrobe](https://github.com/r-lib/usethis/pull/1423/files), which shows how important it's becoming, near other important tools such as withr and rlang.

### What if you were making pretty interfaces with usethis?

Is your package using `usethis::ui_` functions?
If you have time for a bit of upkeep, you can do the switch from that to cli by reading and following the [cli article on the topic](https://cli.r-lib.org/articles/usethis-ui.html).
Do not discover this article *well after* starting a transition, cough (that's what happened to Maëlle).

## cli formatting: all the curly braces

You'll probably end up calling various cli functions whose names start with `cli_` to create [semantic elements](https://cli.r-lib.org/reference/index.html#semantic-cli-elements): headers, lists, alerts, etc.
What do they expect as inputs?

cli has a glue-like syntax: if there's an object called say `thing` and you want to use it in a message, you use `{thing}`.

```{r}
thing <- "a string"
cli::cli_li("Hey you provided this text: {thing}!")
```

Then, you can use classes in messages.
Classes, like classes in CSS, help format the elements.
For instance if `thing` is a variable, we write:

```{r}
thing <- "a string"
cli::cli_li("Hey you provided this text: {.var {thing}}!")
```

See the [full list of classes](https://cli.r-lib.org/reference/inline-markup.html#classes) in cli docs.
To mention only a few of them (you might notice clickability is a star here...if the terminal supports ANSI hyperlinks to runnable code (for instance, RStudio IDE)):

-   `.run` means that the code in the message will be clickable! Best code hints ever!
-   `.help` will have a clickable link to a help topic.
-   `.file` will have a clikable link to a file.
-   `.obj_type_friendly`, for instance `{.obj_type_friendly {mtcars}}`, prints an object in, well, a friendly way (thanks to Jon Harmon for reminding us about this one).

It's well worth going through the list of classes at least once.

### What if my string contains curly braces?

If you want to actually communicate something with curly braces, you'll need to [double them to escape them](https://cli.r-lib.org/reference/inline-markup.html#escaping-and-).

### What about plural?

cli has support for [pluralization](https://cli.r-lib.org/reference/pluralization.html) (presumedly only for 0/1/more than one, not for more complex forms of pluralization).

### How to add a custom class / theme

How to define a custom class or theme seems to be a bit underdocumented at the moment, which is unsurprising as it's an advanced topic.
Jenny Bryan [assumes](https://mastodon.social/@jennybryan@fosstodon.org/110675320334403080) adding a custom class needs to happen as part of a theme, which she does in [googlesheets4](https://github.com/tidyverse/googlesheets4/blob/fdb187643b324cd607f71cefa133bf49924f6e49/R/utils-ui.R#L1-L16).
Michael McCarthy [recommends](https://mastodon.social/@mccarthymg@fosstodon.org/110673498809652527) looking at what cli itself does.

Open questions remain: How to let an user override it with their own?
What about dark themes?

### How to turn off colors

cli respects the [nocolor standard](https://cli.r-lib.org/articles/cli-config-user.html?q=no#no_color), therefore an user can set the environment variable `NO_COLOR` to any string in order to not get colored output.

## cli for expressing what?

### Advertise side effects

See the the [tidy design guide](https://design.tidyverse.org/spooky-action.html?q=cli#advertise-the-side-effects).

### Progress

As seen when installing packages with [pak](https://pak.r-lib.org/)!
Not only does cli support progress bars, it [documents how to create them in two articles](https://cli.r-lib.org/articles/progress.html)!

#### How does it look in logfiles?

The cli output in non-interactive sessions is discussed in the [article about advanced progress bar topics](https://cli.r-lib.org/articles/progress-advanced.html#non-interactive-r-sessions)

### Error messages

You can use `cli::cli_abort()` instead of `stop()` for errors, as recommended by the [tidyverse style guide](https://style.tidyverse.org/error-messages.html).
Notes that using this function necessitates your package importing rlang.

## How to make cli quiet or not

You can choose to silence/shush/muffle cli messages!

For cli functions whose name starts with `ci_`, see the [docs](https://cli.r-lib.org/articles/semantic-cli.html#cli-messages).

Let's try an example.

```{r}
cli::cli_li("hello")
rlang::local_options(cli.default_handler = function(msg) invisible(NULL))
cli::cli_li("hello")
```

This can be useful in tests for instance!
Who wants to be looking at a wall of cli output when debugging one's tests.

For other functions, probably write your own wrapper and make it responsive to an option (rather than arguments in all functions, see [discussion](https://github.com/ropensci/dev_guide/issues/603)), like what happens in usethis with the `usethis.quiet` option.

Speaking of options, let's remind you about `rlang::is_interactive()` whose output is controllable via the `rlang_interactive` global option.

### How to test cli output

You should probably look into [snapshot tests](https://testthat.r-lib.org/articles/snapshotting.html), maybe in combination with [`cli::test_that_cli()`](https://cli.r-lib.org/reference/test_that_cli.html).

## Conclusion

In this post we presented what we know about cli: in short, it's a fantastic package for building informative and pretty command-line interfaces, and its docs are extensive!
