---
slug: mocking-new-take
title: "Update on mocking for testing R packages" 
authors: 
- Maëlle Salmon
date: "2024-03-13" 
tags: 
- package development
- testing
- mocking
output: hugodown::hugo_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.path = "", comment = "")
# knitr hook to make images output use Hugo options
knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)

```

```{r, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  asciicast_theme = "pkgdown"
)
```

This blog featured a [post on mocking](/2019/10/29/mocking/) years ago.
Since then, we've entered a new decade, the second edition of Hadley Wickham's and Jenny Bryan's R packages book was published, and mocking came back to testthat, so it's time for a new take/resources roundup!

## Mocking yay or nay

The R packages book by Hadley Wickham and Jenny Bryan contains an insightful paragraph on [test coverage](https://r-pkgs.org/testing-design.html#sec-testing-design-coverage)[^1], including those lines:

[^1]: The three chapters on testing are definitely worth a read!

> "In many cases, that last 10% or 1% often requires some awkward gymnastics to cover. Sometimes this forces you to introduce mocking or some other new complexity. Don't sacrifice the maintainability of your test suite in the name of covering some weird edge case that hasn't yet proven to be a problem."

This is definitely good to keep in mind whilst diving into the world of mocking.
Not that mocking needs to be especially complex!

Beside, sometimes, testing is complex because your code is begging to be refactored.
:wink:

## Revisiting the general mocking example

The [example of general mocking](/2019/10/29/mocking/#general-mocking) in the post from 2019 can be rewritten to take advantage of the new mocking functionality that was [announced as no longer experimental in October 2023](https://www.tidyverse.org/blog/2023/10/testthat-3-2-0/#mocking).

Now, using testthat's mocking means the best demo includes an actual package... which thankfully can be created in a few usethis function calls, and writing a few code lines (well, copy-pasting them from the old post!).

``` r
create_package("../mockexample")
# then from mockexample
use_mit_license()
use_r("encoding")
use_testthat()
use_test("encoding")
# then filling in the script and test file
use_git()
use_github()
# then remember it'd be nice to show it works
use_github_action("check-standard")
# Git add, commit and push
```

Find the example repository on [GitHub](https://github.com/maelle/mockexample)

The mockexample package has a function called `is_encoding_a_pain()`:

``` r
is_encoding_a_pain <- function(){
  if (Sys.info()[["sysname"]] == "Windows"){
    return(TRUE)
  }

  FALSE
}
```

The function returns `TRUE` on Windows, `FALSE` on other operating systems.

To test it no matter the operating system, we need to make our tests believe they are run on a given operating system.
We can in this day and age achieve this using mocking tools from testthat itself!
Here's our test file:

``` r
test_that("is_encoding_a_pain() works on Linux", {
  local_mocked_bindings(Sys.info = function(...) c(sysname = "Linux"))
  expect_false(is_encoding_a_pain())
})

test_that("is_encoding_a_pain() works on Windows", {
  local_mocked_bindings(Sys.info = function(...) c(sysname = "Windows"))
  expect_true(is_encoding_a_pain())
})
```

The magic happens in `local_mocked_bindings()` calls such as `local_mocked_bindings(Sys.info = function(...) c(sysname = "Windows"))`, that indicates what we want the mocked function, `Sys.info()`, to return in this environment.
Using testthat's own mocking feels good for three reasons: not needing to add another dependency (and to remember what the mocking packages are called :sweat_smile:), knowing this mocking implementation is [considered better](https://www.tidyverse.org/blog/2023/10/testthat-3-2-0/#mocking) by testthat authors, and getting a [withr vibe](https://withr.r-lib.org/) from the "local\_" aspect.

Now, there's a catch here because `Sys.info()` is a base R function.
We had to add this line in the `R/encoding.R` script:

``` r
Sys.info <- NULL
```

This case is documented, as other cases (namespaced calls in particular), in [testthat::local_mocked_bindings()\` manual page](https://testthat.r-lib.org/reference/local_mocked_bindings.html#use).

TODO: example from keyring?

## Revisiting the escape hatch example

Even the example in a more recent post, on [escape hatches](/2023/01/23/code-switch-escape-hatch-test/), can be easily re-written to use mocking.
It does not completely throw the escape hatch pattern out of the window, but seriously limits its relevance.
