---
slug: input-checking
title: "Checking the inputs of your R functions" 
authors: 
- Sam Abbott
- Hugo Gruson
- Carl Pearson
- Tim Taylor
date: "2021-10-07" 
tags: 
- package development 
- r-package
output: hugodown::hugo_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.path = "", comment = "")
# knitr hook to make images output use Hugo options
knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)
options(crayon.enabled = FALSE)

```

## Introduction: the dangers of not checking function inputs

R functions and R packages are a convenient way to share code with the rest of the world but it is generally not possible to know how, or with what precise aim in mind, others will use your code.
For example, they might try to use it on objects that your function was not designed for.
Let's imagine we have written a short function to compute the geometric mean:

```{r}
geometric_mean <- function(...) {
  
  return(prod(...)^(1/...length()))
  
}
```

When you tested the function yourself, everything seemed fine:

```{r}
geometric_mean(2, 8)

geometric_mean(4, 1, 1/32)
```

But a different person using your function might expose it to situations it was not prepared to handle, resulting in cryptic errors or undefined behaviour:

```{r, error = TRUE}
# Input with factors instead of numerics
geometric_mean(factor(2), 8)

# Input with negative values
geometric_mean(-1, 5)

# Input with NAs
geometric_mean(2, 8, NA)
```

Or worse, it could give an incorrect output:

```{r}
geometric_mean(c(2, 8))
```

Because of this, you need to make sure you return clear errors whenever your functions receives input it was not designed for.
In this blog post, we review a range of approaches to help you check your function inputs and discuss some potential future developments.

## Pre-requisite: thoroughly document your argument types

You can notice from the simple example above that it's easy to pass invalid inputs to the `geometric_mean()` function because we didn't provide any documentation on what is or isn't a valid input.
We won't go into details here but the [roxygen2](https://roxygen2.r-lib.org/) package provides a convenient way to generate documentation for R functions.
Try to be as precise as possible when describing the required format for your inputs [^1].

[^1]: [Some package developers even developed their own standardized way to document argument types and length](https://github.com/r-lib/withr/commit/42e503092046705f30032cb3a321d64b0e9383d4).
    But there is currently no standard shared across the R community.

```{r}
#' @param name A character of length one with the name of the person to greet
say_hello <- function(name) {
  stopifnot(is.character(name))
  paste("Hello", name)
}
```

Adding any kind of argument checking in the absence of good documentation would be vain and very frustrating for your users as they would have to figure out what is or isn't valid by trial and error.

## Checking function inputs using base R [^2]

[^2]: Note that these base functions have equivalent in the tidyverse with a more consistent design and coloured output.
    `match.arg()`'s equivalent is `rlang::arg_match()` and `stopifnot()`'s

### `match.arg()`

If the input can only take specific values, the base function `match.arg()` can also prove useful:

```{r, error = TRUE}
match.arg(arg = "R", choices = c("R", "python"))

match.arg(arg = "javascript", choices = c("R", "python"))
```

But the real power of the `match.arg()` function comes from the fact that `choices` can be automatically obtained in the context of a function:

```{r, error = TRUE}
choose_language <- function(language = c("R", "python")) {
  
  # Equivalent to `match.arg(language, c("R", "python"))
  match.arg(language)
  
  paste("I love", language)
  
}

choose_language("R")

choose_language("julia")
```

### `stopifnot()`

There is a another, more general, built-in mechanism to check input values in base R: `stopifnot()`.
You can see it [used](https://github.com/wch/r-source/blob/79298c499218846d14500255efd622b5021c10ec/src/library/stats/R/approx.R#L78) [throughout](https://github.com/wch/r-source/blob/79298c499218846d14500255efd622b5021c10ec/src/library/stats/R/cor.R#L36) [R](https://github.com/wch/r-source/blob/79298c499218846d14500255efd622b5021c10ec/src/library/graphics/R/smoothScatter.R#L47) [source](https://github.com/wch/r-source/blob/79298c499218846d14500255efd622b5021c10ec/src/library/base/R/srcfile.R#L23) [code](https://github.com/wch/r-source/blob/79298c499218846d14500255efd622b5021c10ec/src/library/base/R/parse.R#L65).
As its name suggests, it will *stop* the function execution *if* an object does *not* pass some tests.

```{r, error = TRUE}
say_hello <- function(name) {
  stopifnot(is.character(name))
  paste("Hello", name)
}

say_hello("Bob")
say_hello(404)
```

However, as you can see in this example, the error message is not in plain English but contains some code instructions.
This can hinder understanding of the issue.

Because of this, `stopifnot()` was improved in R 4.0.0:

> stopifnot() now allows customizing error messages via argument names, thanks to a patch proposal by Neal Fultz in PR#17688.

This means we can now provide a clearer error message directly in `stopifnot()` [^3]:

[^3]: Read [the tidyverse style guide](https://style.tidyverse.org/error-messages.html) for more guidance on how to write good error messages.

```{r, error = TRUE}
say_hello <- function(name) {
  stopifnot("`name` must be a character." = is.character(name))
  paste("Hello", name)
}

say_hello(404)
```

This is clearly a really great improvement to the functionality of base R.
However, we can see from this example that we could create the error message programmatically based on the contents of the test.
Each time we test if the object is of `class_X` and this is not true, we could throw an error saying something like "x must of a class_X".
This way, you don't have to repeat yourself which is generally a good aim [^4].
This becomes necessary when you start having many input checks in your function or in your package.

[^4]: The [Don't Repeat Yourself (DRY) principle of software development](https://en.wikipedia.org/wiki/Don't_repeat_yourself), also mentioned in this post on [caching](https://blog.r-hub.io/2021/07/30/cache/)

## Checking function inputs using R packages

### The example of the checkmate package

Although some developers create [their own functions](https://github.com/djnavarro/bs4cards/blob/a021d731a307ec7af692a42364308b60e2bf9827/R/validators.R) to solve this problem, you can also rely on existing packages to make your life easier.
One of these packages designed to help you in input checking is [checkmate](https://mllg.github.io/checkmate/).
checkmate provides a large number of functions that check that inputs respect a given set of properties, and that return clear error messages when that is not the case:

```{r}
say_hello <- function(name) {
  # Among other things, check_string() checks that we provide a 
  # character object of length one
  checkmate::assert_string(name)
  paste("Hello", name)
}
```

```{r, error = TRUE}
say_hello(404)
```

```{r, error = TRUE}
say_hello(c("Bob", "Alice"))
```

### Other packages to check function inputs

Because input checking is such an important point task and because it is so difficult to get right, it is not surprising that there are many packages other than checkmate to solve this issue.
We will not get into the details of all of the available options here but below is a list of some of them, listed by decreasing number of reverse dependencies.
If interested in understanding the various approaches to input taking the documentation for these package is a great place to start.

-   [assertthat](https://github.com/hadley/assertthat)

```{r, error = TRUE}
assertthat::assert_that(is.character(1))
```

-   [assertr](https://docs.ropensci.org/assertr/)

```{r, error = TRUE}
library(magrittr)

mtcars %>%
  assertr::verify(nrow(.) < 10)
```

-   [assertive](https://bitbucket.org/richierocks/assertive)

```{r, error = TRUE}
assertive::assert_is_a_string(1)
```

-   [ensurer](https://github.com/smbache/ensurer)

```{r, error = TRUE}
ensure_square <- ensurer::ensures_that(NCOL(.) == NROW(.))

ensure_square(matrix(1:20, 4, 5))
```

-   `vctrs::vec_assert()`

```{r, error = TRUE}
vctrs::vec_assert(c(1, 2), "character")

vctrs::vec_assert(c(1, 2), size = 3)
```

-   [check](https://github.com/moodymudskipper/check) is slightly different because it doesn't provide utilities that work out of the box, but rather tools to assist you in writing your own checking functions

```{r, error = TRUE}
library(check)

check::setup() 

set_check_fun(
  "`{var}` must be a {type} vector of length {length}." = {
      val <- get(var, env)
      is.atomic(val) && is(val, type) && length(val) == length
  }
)

say_hello <- function(name) {
  check(
    "`name` must be a character vector of length 1."
    )
  paste("hello", name)
}

say_hello("Maria")

say_hello(c("Maria", "Noelia"))
```

## What about the future?

In this post, we have discussed some methods to check function inputs, and to generate more informative error messages when doing so.
However, this always comes with a performance cost, even though it's often relatively limited.
Zero-cost assertions, as found in some other languages, would require some kind of typing system which R does not currently support.
Interestingly several other languages have evolved to have typing systems as they have developed.
Typescript developed as an extension of JavaScript, and type annotations are now possible in Python.
[Will R one day follow suit?](https://blog.q-lang.org/posts/2021-10-16-project/)
