---
title: rversions 1.0.4 is on CRAN!
date: '2019-04-15'
slug: rversions-1-0-4
---

Version 1.0.4 of the `rversions` package has been released on CRAN! `rversions` queries the main R SVN repository to find the versions r-release and r-oldrel refer to, and also all previous R versions and their release dates.

Get `rversions`'s latest version from 
a [safe](https://rud.is/b/2019/03/03/cran-mirror-security/) CRAN mirror near
you:

```r
install.packages("rversions")
```

This release, compared to `rversions` 1.0.3, brings :sparkles: caching :sparkles: of the R versions within each R session (less useless queries!) thanks to a contribution by [Rich FitzJohn](https://github.com/richfitz); a documentation website built with `pkgdown`; and a new home for `rversions`' source from METACRAN to R-hub! :house: Not too much, but since it's the first time we blog about `rversions`, we'll use the release as an opportunity to introduce it with all due respect! How does `rversions` work and what is it useful for?

# Where does `rversions` get its information from?

As you might have guessed, R is developed under version control. The R Core team uses SVN, i.e.[Apache Subversion](https://en.wikipedia.org/wiki/Apache_Subversion). The versions `rversions` guesses are derived from SVN _tags_ in the [main R SVN repository](https://svn.r-project.org/R/). In SVN lingo, ["tagging refers to labeling the repository at a certain point in time so that it can be easily found in the future."](https://en.wikipedia.org/wiki/Apache_Subversion#Branching_and_tagging) so it might remind you of GitHub releases. In more details,

- [r-versions are coming from SVN tags](https://github.com/metacran/rversions/blob/36deaf5c4ffd648c7524d8e3225871e6d97d9ccf/R/rversions.R#L79)

```{r r-versions}
head(rversions::r_versions())
```

- r-release is the latest SVN tag

```{r r-release}
rversions::r_release()
```

- r-oldrel is the latest of the previous minor version

```{r r-oldrel}
rversions::r_oldrel()
```

- r-release-tarball is the latest SVN tag that has a corresponding tar.gz to download

```{r r-release-tarball}
rversions::r_release_tarball()
```

- Windows, macOS are the same: latest SVN tag that has a file to download

```{r win-r}
rversions::r_release_win()
rversions::r_release_macos()

```

# Where is `rversions` used?

`rversions` is used in R packages but also public web services for R package developers, including R-hub of course!

### In R packages

* `rversions` is used in `devtools`, in [`dr_devtools()`](https://devtools.r-lib.org/reference/dr_devtools.html) to compare the installed R version to the current R release version.

* [`provisionr`](https://github.com/mrc-ide/provisionr) imports `rversions` and uses it in `check_r_version()`, a function that checks and coerces something into an R version.

### In other tools

- [r-appveyor web service uses `rversions` to determine what versions r-release and r-oldrel are](https://github.com/krlmlr/r-appveyor/blob/master/scripts/appveyor-tool.ps1#L77-L87).

- R-hub uses it to do the same on [Solaris](https://github.com/r-hub/solarischeck/blob/d3edc6077f52081a53b2ed8b9ec40e17176fdb43/run.sh#L105), [macOS](https://github.com/r-hub/macoscheck/blob/846f87d1ad1c4b4aeb283a986e084913f203879f/run.sh#L160) and [Windows](https://github.com/r-hub/wincheck/blob/591ea8d9a28cd5044c063e88f46ef0507fa16fc7/run.ps1#L114).

- [R-hub's Windows install script uses it as well, to decide what versions of R to install](https://github.com/r-hub/rhub-server/blob/master/jenkins/scripts/windows-2008-setup.ps1#L74-L79).


# Can we also _play_ with `rversions` data?

At this point, you have no doubt `rversions` is useful, but what about we use it to get a glimpse at R's release schedule over time?


```{r}
library("ggplot2")
versions <- rversions::r_versions()

```

# Conclusion

The best place to get to know `rversions` is [its brand-new documentation website built with `pkgdown`](https://r-hub.github.io/rversions), and the best place to provide feedback or contribute is [its GitHub repo](https://github.com/r-hub/rversions). Thanks to all folks who chimed in in the issue tracker and pull request tab!
