---
slug: lazy-meanings
title: "Lazy introduction to laziness in R" 
authors: 
- Maëlle Salmon
- Athanasia Mo Mowinckel
date: "2025-02-01" 
tags: 
- package development
- programming
output: hugodown::hugo_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.path = "", comment = "")
# knitr hook to make images output use Hugo options
knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)

```

```{r, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  asciicast_theme = "pkgdown"
)
```

In the programming world, laziness can often be a good thing: it is both a human quality that can motivate automation efforts, and a programming concept that avoids wasting resources such as memory.
Now, when reading code or documentation, seeing the word "lazy" can be confusing, because of its polisemy: it carries several meanings.
In this post, we will enumerate the different possible definitions of "lazy" in R code.

## Lazy as in lazy evaluation

You might know that R provides **lazy evaluation**: the arguments of a function are only evaluated if they are accessed.
In short, you can pass anything as an argument value to a function without any problem as long as the function does not use that value.

For instance, the code below works despite `evaluation`'s not existing, because the definition of the `mean()` function includes ellipsis, and because the `lazy` argument is actually not used.

```{r}
mean(1:10, lazy = evaluation)
```

The contrary of lazy evaluation is **eager evaluation**.

The [Advanced R book by Hadley Wickham](https://adv-r.hadley.nz/functions.html#lazy-evaluation) features a very clear introduction to lazy evaluation.

Note that the workhorse of lazy evaluation in base R is a thing called a **promise** that contains an *expression* (the recipe for getting a value), an *environment* (the ingredients that are around), and a *value*.
The latter is only computed when accessed, and cached once computed.

### What about {future}?

Maybe you have heard of the [future package](https://future.futureverse.org/index.html) by Henrik Bengtsson.
It provides an implementation in R of **futures**, a programming concept.
Its homepage state "In programming, a future is an abstraction for a value that may be available at some point in the future. The state of a future can either be unresolved or resolved."

With futures, you create a future, that is associated to a **promise**, which is a **placeholder for a value** and then the value itself (so not the same definition of "promise" as the "promises" used by base R in the context of lazy evaluation).
The value can be computed asynchronously, which means in parallel.
Therefore, the futures package allows R programmers to take full advantage of their local computing resources: cores, clusters, etc.

Now, to add confusion beyond the different meaning of "promise" in this context, [by default](https://future.futureverse.org/reference/future.html) a future is **not lazy**, it is **eager**.
This means that it is computed immediately.

By default, the creation of a future below (`eager_future`) takes as much time as not wrapping the code in a future, because the computation is immediate.
Setting `lazy` to `TRUE` makes the future creation much faster (`lazy_future`).

```{r, cache=TRUE}
library("future")
bench::mark(
  no_future = is.numeric(runif(n = 10000000)),
  eager_future = future(is.numeric(runif(n = 10000000))),
  lazy_future = future(is.numeric(runif(n = 10000000)), lazy = TRUE),
  check = FALSE
)
```

If we do retrieve the value, overall the same time is spent between creating the future and our getting the value:

```{r, cache=TRUE, warning=FALSE}
bench::mark(
  no_future = {is.numeric(runif(n = 10000000))},
  eager_future = {x <- future(is.numeric(runif(n = 10000000))); value(x)},
  lazy_future = {x <- future(is.numeric(runif(n = 10000000)), lazy = TRUE); value(x)},
  check = FALSE
)
```

Therefore, the use of futures and the use of lazy evaluation are orthogonal concepts: you can use future with or without lazy evaluation.
The future package is about *how* the value is computed (in parallel or sequentially for instance), lazy evaluation is about *when* the value is computed (right as it is defined, or only when it is needed).

## Lazy as in lazy database operations

In the database world, queries can be lazy: the query is like a TODO list that is only executed (computed, evaluated) when you want to access the resulting table or result.
Making the output tangible is called *materialization*.

This is vocabulary we can encounter when using:

-   the [dbplyr package](https://dbplyr.tidyverse.org/), which is the dplyr back-end for databases. "All dplyr calls are evaluated lazily, generating SQL that is only sent to the database when you request the data."

Slightly tweaked from the dbplyr README,

```{r}
# load packages
library(dplyr, warn.conflicts = FALSE)
library("dbplyr")

# create the connection and refer to the table
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
copy_to(con, mtcars)
mtcars2 <- tbl(con, "mtcars")

# create the query
summary <- mtcars2 %>% 
  group_by(cyl) %>% 
  summarise(mpg = mean(mpg, na.rm = TRUE)) %>% 
  arrange(desc(mpg))

# the object is lazy, the value is not computed yet
summary
nrow(summary)

# we explictly request the data, so now it's there
answer <- collect(summary)
nrow(answer)

```

-   the [dtplyr package](https://dtplyr.tidyverse.org/index.html), which is a data.table back-end for dplyr. The ["lazy" data.table objects](https://dtplyr.tidyverse.org/reference/lazy_dt.html) "captures the intent of dplyr verbs, only actually performing computation when requested" (with `collect()` for instance). The manual also explains that this allows dtplyr to make the code more performant by simplifying the data.table calls.

Slightly tweaked from dtplyr README,

```{r}
# load packages
library(data.table)
library(dtplyr)
library(dplyr, warn.conflicts = FALSE)

# create a “lazy” data table that tracks the operations performed on it.
mtcars2 <- lazy_dt(mtcars)

# create the query
summary <- mtcars2 %>% 
  filter(wt < 5) %>% 
  mutate(l100k = 235.21 / mpg) %>% # liters / 100 km
  group_by(cyl) %>% 
  summarise(l100k = mean(l100k))

# the object is lazy, the value is not computed yet
summary
nrow(summary)

# we explictly request the data, so now it's there
answer <- as_tibble(summary)
nrow(answer)

```

-   the [duckplyr package](https://duckplyr.tidyverse.org/dev/), which is a drop-in replacement for dplyr, powered by DuckDB for fast operation. Objects created with `duckdb_tibble(.lazy = TRUE)` (`.lazy = TRUE` is not the default) and with `read_parquet_duckdb()` (`lazy=TRUE` is the default) are lazy, the value is only computed if explicitly accessed.

### duckplyr, lazy evaluation and deferred evaluation

Some doubts to be clarified

## Lazy as in lazy loading of data in packages (`LazyData`)

If your R package exports data, and sets the `LazyData` field in `DESCRIPTION` to `true`, then the exported datasets are lazily loaded: they're available without the use of `data()`, but they're not actually taking up memory until they are accessed.

There's more details on `LazyData` in the [R packages book by Hadley Wickham and Jenny Bryan](https://r-pkgs.org/data.html#sec-data-data) and in [Writing R Extensions](https://cloud.r-project.org/doc/manuals/r-devel/R-exts.html#Data-in-packages).

Note that internal data is always lazily loaded, and that data that is too big cannot be lazily loaded.

## Lazy as in frugal file modifications

The `pkgdown::build_site()` function, that creates a documentation website for an R package, features a [`lazy` argument](https://pkgdown.r-lib.org/reference/build_site.html#arg-lazy).
"If `TRUE`, will only rebuild articles and reference pages if the source is newer than the destination."

It is a much simpler concept of laziness: decide right now whether it is needed to rebuild each page.

The potools package, that provides tools for portability and internationalization of R packages, uses ["lazy" for a similar meaning](https://michaelchirico.github.io/potools/reference/po_update.html?q=lazy#ref-usage).

## Lazy as in lazy quantifiers in regular expressions

In regular expressions you can use [quantifiers](https://blog.djnavarro.net/posts/2024-12-16_regex-backreferences/#quantifiers) to indicate how many times a pattern must appear: the pattern can be optional, appear several times, etc.
You can also specify whether the tool should match as many repetitions as possible, or the fewest number of repetitions possible.

Matching the fewest number of repetitions possible is "lazy" (stingy, I made that up).
Matching as many repetitions as possible is "eager" (or greedy, actual term).

```{r}
string <- "aaaaaa"
# greedy! eager!
stringr::str_match(string, "a+")
# stingy! lazy!
stringr::str_match(string, "a+?")
```

## Conclusion

In the context of lazy evaluation and lazy database operations we can think of lazy as a sort of parcimonious procrastination.
In the case of frugal file modifications in pkgdown and potools, lazy means an informed decision is made on the spot on whether a computation is needed.
In the case of lazy quantifiers in regular expressions, lazy means stingy.

Overall, an user can expect "lazy" to mean "less waste", but it is crucial that the documentation of the particular piece of software at hand clarifies the meaning and the potential trade-offs.
